The Gillespie Algorithm: Stochastic Modeling of Molecular Reaction Networks

### 2024-11-28 paper/paper.Rmd
clean files

---
title: "The Gillespie Algorithm: Stochastic Modeling of Molecular Reaction Networks"
author: 
  - Jiayu Guo
  - Sinan Ma 
  - Anna Szydlowski
date: "`r Sys.Date()`"
abstract: "The Gillespie Algorithm provides a robust framework for simulating stochastic molecular reaction networks, capturing the inherent randomness of chemical and biochemical processes. Unlike deterministic methods, the algorithm accurately models systems with low molecular counts, where fluctuations significantly influence reaction dynamics. This project explores the theoretical foundation of the Gillespie Algorithm, including its basis in the master equation, and demonstrates its application in simulating two molecular systems: the photodimerization of thymine, a process linked to UV-induced DNA damage, and the radioactive decay of isotopes. Through these simulations, the algorithm's capability to model stochastic trajectories is highlighted, offering insights into reaction mechanisms and the role of randomness in molecular processes. The implications of these findings extend to a variety of disciplines, including molecular biology, environmental monitoring, and nuclear medicine."
output:
  pdf_document:
    number_sections: true
    citation_package: biblatex
linkcolor: blue
urlcolor: blue
spacing: double
header-includes:
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyhead[L]{The Gillespie Algorithm}
  - \fancyfoot[CO,CE]{}
  - \fancyfoot[LE,RO]{\thepage}
  - \usepackage{setspace}
  - \doublespacing

@@

#### General Explanation of Exponential Decay
The radioactive decay of \( R \) follows the well-known exponential decay law:

@@

#### Analysis of Figure \ref{fig:decay-linear}
Figure \ref{fig:decay-linear} illustrates the exponential decay of \( R \) over time. The steep initial decline in the curve corresponds to the high decay rate at the beginning of the simulation when \( R \) is at its maximum. As \( R \) decreases, the curve flattens, reflecting the slower decay rate. This behavior exemplifies the exponential decay dynamics, where the number of remaining molecules reduces rapidly at first and progressively lessens as the substance depletes.

@@

\newpage
#### Analysis of Figure \ref{fig:decay-log}
 Figure \ref{fig:decay-log} depicts the same decay trajectory on a logarithmic scale, offering a complementary perspective. On this scale, the exponential decay law is linearized, and the trajectory appears as a straight line. The slope of the line corresponds to the negative decay constant, \(-c\), providing a quantitative validation of the theoretical model.

@@

#### Radioactive Half-Life
The exponential decay observed in the simulation aligns with the concept of half-life, a critical measure in the study of radioactive processes. The half-life (\( t_{1/2} \)) is defined as the time required for half of the radioactive substance to decay (@Henriksen2002). It can be calculated as:

@@

#### Implications and Real-World Applications
Radioactive decay is a fundamental phenomenon in nuclear physics and has wide-ranging applications across multiple scientific disciplines. For instance, in nuclear medicine, the controlled decay of radioactive isotopes is utilized in radiotherapy to deliver targeted doses of radiation for cancer treatment. Similarly, in environmental science, the decay of isotopes such as radon is monitored to assess contamination levels and potential health risks.

The stochastic approach modeled here provides critical insights into systems with low molecule counts, where randomness plays a significant role. For example, in radiological dosimetry, small-scale decay dynamics can impact the precision of dose delivery. Likewise, in environmental monitoring, understanding the stochastic nature of decay processes aids in the interpretation of measurement data, particularly in situations where low concentrations of isotopes are involved.

@@

# Conclusion {#sec-conclusion}
The Gillespie Algorithm represents a significant advancement in the numerical simulation of stochastic molecular reaction networks. Its ability to generate statistically valid trajectories of reaction systems provides critical insights into processes governed by inherent randomness, especially in systems with low molecule counts. This study has illustrated the versatility and power of the algorithm through two applications: the photodimerization of thymine and the radioactive decay of isotopes.

The simulation of thymine photodimerization highlighted the algorithm’s capacity to model complex reaction networks, offering a detailed view of the formation of UV-induced DNA lesions such as CPDs and 64-PPs. This has direct implications for understanding the molecular origins of UV-induced mutagenesis and skin cancer. Similarly, the radioactive decay simulation demonstrated the algorithm’s precision in modeling exponential decay processes and capturing stochastic fluctuations, a crucial feature for understanding radioactive isotopes in fields such as nuclear medicine and environmental science.

@@

\appendix
\section{Appendix}
## R Code for [4.1 Photodimerization Simulation](#sec-Photodimerization)

```{r a1, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
# Load required libraries
if (!require("reshape2")) install.packages("reshape2")
if (!require("ggplot2")) install.packages("ggplot2")
library(reshape2)
library(ggplot2)
# Define reaction rates
c1 <- (37 / 1e6) * 10             # Formation of CPD
c2 <- (2 / 1e6) * 10              # Formation of 64-PP
# Initial state
state <- c(T1 = 10000, T2 = 10000, CPD = 0, `64PP` = 0)
# Define propensity functions
propensities <- function(state) {
  c(
    c1 * state["T1"] * state["T2"], # Reaction 1
    c2 * state["T1"] * state["T2"]  # Reaction 2
  )
}
# Gillespie simulation function
simulate_gillespie <- function(state, max_time) {
  # Initialize time and output data frame
  time <- 0
  output <- data.frame(
    Time = numeric(0),
    T1 = numeric(0),
    T2 = numeric(0),
    CPD = numeric(0),
    `64PP` = numeric(0)
  )

@@

  # Convert columns to numeric for consistency
  output[] <- lapply(output, as.numeric)
  return(output)
}
# Run the simulation
max_time <- 100
simulation <- simulate_gillespie(state, max_time)

@@

## R code for [4.2 Radioactive Decay Simulation](#sec-Decay)
```{r a2, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
# Define parameters
c <- 0.01  # Decay rate constant
state <- c(R = 10000)  # Initial radioactive substance
max_time <- 1000  # Extended maximum simulation time

@@

# Linear-scale plot
ggplot(simulation, aes(x = Time, y = R)) +
  geom_line(color = "blue") +
  labs(
    x = "Time",
    y = "Remaining Radioactive Substance (R)"
  ) +
  theme_minimal()

# Logarithmic-scale plot
ggplot(simulation, aes(x = Time, y = R)) +
  geom_line(color = "red") +
  scale_y_log10() +
  labs(
    x = "Time",
    y = "Remaining Radioactive Substance (log(R))"
  ) +
  theme_minimal()

>>>>

---
title: "The Gillespie Algorithm: Stochastic Modeling of Molecular Reaction Networks"
author: 
  - Jiayu Guo
  - Sinan Ma 
  - Anna Szydlowski
thanks: "We would like to thank Dr. Omidali A. Jazifor for his invaluable guidance throughout this project. The source code for simulations can be found on our [GitHub repository](https://github.com/masinan123/Gillespie-algorithm.git)."
date: "`r Sys.Date()`"
abstract: "The Gillespie Algorithm provides a framework for simulating stochastic molecular reaction networks, capturing the inherent randomness of chemical and biochemical processes. Unlike deterministic methods, the algorithm accurately models systems with low molecular counts, where fluctuations significantly influence reaction dynamics. This project explores the theoretical foundation of the Gillespie Algorithm, including its basis in the master equation, and demonstrates its application in simulating two molecular systems: the photodimerization of thymine, a process linked to UV-induced DNA damage, and the radioactive decay of isotopes. Through these simulations, the algorithm's capability to model stochastic trajectories is highlighted, offering insights into reaction mechanisms and the role of randomness in molecular processes. The implications of these findings extend to a variety of disciplines, including molecular biology, environmental monitoring, and nuclear medicine."
output:
  pdf_document:
    number_sections: true
    citation_package: biblatex
linkcolor: blue
urlcolor: blue
spacing: double
header-includes:
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyhead[L]{The Gillespie Algorithm}
  - \fancyfoot[CO,CE]{}
  - \fancyfoot[LE,RO]{\thepage}
  - \usepackage{setspace}
  - \doublespacing

@@

#### General Explanation of Exponential Decay

The radioactive decay of \( R \) follows the well-known exponential decay law:

@@

#### Analysis of Radioactive Decay Simulation in Linear Scale (Figure \ref{fig:decay-linear})

Figure \ref{fig:decay-linear} illustrates the exponential decay of \( R \) over time. The steep initial decline in the curve corresponds to the high decay rate at the beginning of the simulation when \( R \) is at its maximum. As \( R \) decreases, the curve flattens, reflecting the slower decay rate. This behavior exemplifies the exponential decay dynamics, where the number of remaining molecules reduces rapidly at first and progressively lessens as the substance depletes.

@@

\newpage
#### Analysis of Radioactive Decay Simulation in Logarithmic Scale (Figure \ref{fig:decay-log})

Figure \ref{fig:decay-log} depicts the same decay trajectory on a logarithmic scale, offering a complementary perspective. On this scale, the exponential decay law is linearized, and the trajectory appears as a straight line. The slope of the line corresponds to the negative decay constant, \(-c\), providing a quantitative validation of the theoretical model.

@@

#### Radioactive Half-Life

The exponential decay observed in the simulation aligns with the concept of half-life, a critical measure in the study of radioactive processes. The half-life (\( t_{1/2} \)) is defined as the time required for half of the radioactive substance to decay (@Henriksen2002). It can be calculated as:

@@

#### Implications and Real-World Applications

Radioactive decay is a fundamental phenomenon in nuclear physics and has wide-ranging applications across multiple scientific disciplines. For instance, in nuclear medicine, the controlled decay of radioactive isotopes is utilized in radiotherapy to deliver targeted doses of radiation for cancer treatment (@britannica_radioactivity). Similarly, in environmental science, the decay of isotopes such as carbon-14 is monitored to assess contamination levels and potential health risks (@openmedscience_carbon14_decay).

The stochastic approach modeled here provides insights into systems with low molecule counts, where randomness plays a significant role. For example, in radiological dosimetry, small-scale decay dynamics can impact the precision of dose delivery. In environmental monitoring, understanding the stochastic nature of decay processes aids in the interpretation of measurement data, particularly in situations where low concentrations of isotopes are involved.

@@

# Conclusion {#sec-conclusion}

The Gillespie Algorithm represents a significant role in the numerical simulation of stochastic molecular reaction networks. Its ability to generate statistically trajectories of reaction systems provides insights into processes governed by inherent randomness, especially in systems with low molecule counts. This project has illustrated the versatility and power of the algorithm through two applications: the photodimerization of thymine and the radioactive decay of isotopes.

The [thymine photodimerization simulation](#sec-Photodimerization) illustrated the algorithm’s capacity to model complex reaction networks, offering a detailed view of the formation of UV-induced DNA lesions such as CPDs and 64-PPs. This has direct implications for understanding the molecular origins of UV-induced mutagenesis and skin cancer. Similarly, the [radioactive decay simulation](#sec-Decay) demonstrated the algorithm’s precision in modeling exponential decay processes and capturing stochastic fluctuations, a crucial feature for understanding radioactive isotopes in fields such as nuclear medicine and environmental science.

@@

\appendix
\section{Appendix}
## R Code for [Photodimerization Simulation](#sec-Photodimerization) (section 4.1)

### Simulation

```{r application-1, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
# Work environment set-up
library(reshape2)
library(ggplot2)
# Define reaction rates
c1 <- (37 / 1e6) * 10             # Formation of CPD
c2 <- (2 / 1e6) * 10              # Formation of 64-PP
# Initial state
state <- c(T1 = 10000, T2 = 10000, CPD = 0, `64PP` = 0)
# Define propensity functions
propensities <- function(state) {
  c(
    c1 * state["T1"] * state["T2"], # Reaction 1
    c2 * state["T1"] * state["T2"]  # Reaction 2
  )
}
# Gillespie simulation function
simulate_gillespie <- function(state, max_time) {
  # Initialize time and output data frame
  time <- 0
  output <- data.frame(
    Time = numeric(0),
    T1 = numeric(0),
    T2 = numeric(0),
    CPD = numeric(0),
    `64PP` = numeric(0)
  )

@@

  # Convert columns to numeric for consistency
  output[] <- lapply(output, as.numeric)
  return(output)
}
```

### Visulization for Figure \ref{fig:dimerization}

```{r figure-1, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
# Run the simulation
max_time <- 100
simulation <- simulate_gillespie(state, max_time)

@@

## R code for [Radioactive Decay Simulation](#sec-Decay) (section 4.2)

### Simulation

```{r application-2, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
# Define parameters
c <- 0.01  # Decay rate constant
state <- c(R = 10000)  # Initial radioactive substance
max_time <- 1000

@@

# Run the simulation
simulation <- simulate_gillespie(state, max_time)
```

### Visulization for Figure \ref{fig:decay-linear}

```{r figure-2, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
# Load ggplot2 for visualization
library(ggplot2)

### Visulization for Figure \ref{fig:decay-log}

```{r figure-3, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE, include=TRUE}
# Logarithmic-scale plot
ggplot(simulation, aes(x = Time, y = R)) +
  geom_line(color = "red") +
  scale_y_log10() +
  labs(
    x = "Time",
    y = "Remaining Radioactive Substance (log(R))"
  ) +
  theme_minimal()
